Self-assembly of colloid-cholesteric composites: A route to switchable optical materials

### 2013-10-15 trunk/papers/2012-sandwich/support.tex
updated supporting material
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2167 40876441-3496-0410-a613-a2fca35b2ae2
updated supporting material
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2169 40876441-3496-0410-a613-a2fca35b2ae2
Added structure factor plots and defintion
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2178 40876441-3496-0410-a613-a2fca35b2ae2

In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
free energy density is also present to represent the energetic cost of
anchoring at a solid surface. If the preferred order parameter tensor
at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
density (per unit area) is
\begin{equation}
f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
\end{equation}
where $W$ is a constant determining the strength of the anchoring.
The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
discussed below. For a colloid of radius $a$, the strength of the surface
anchoring compared which the bulk fluid elastic constant may be quantified
by the dimensionless parameter $Wa/K$. For small values of this parameter,
the presence of a colloid surface should have little impact on the local
fluid LC ordering.

@@

The surface anchoring for colloids is as before: always normal, but with
$Wa/K = 0.23$ and $Wa/K = 2.3$. The colloid solid volume fractions are
1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
in simulation units. All simulations are run for two million simulation
time steps.

@@

The following protocol is used for the simulations of BPIII with
colloidal particles. An initial configuration of
randomly oriented and positioned double twist cylinders is used,
from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
is equilibrated for $6\times10^{5}$ LB time steps until no significant
further evolution is observed. The colloidal particles, with weak
normal anchoring, are then inserted with random chosen positions,
and the composite system is equilibrated for another $1.2\times10^{6}$
time steps.
At the end of the equilibration phase the uniform electric field
with reduced field strength ${\cal E} = 0.8$ 
is switched on (oriented along the $z$-direction), and the simulation
run for a further $1.2\times10^{6}$ time steps.
This is found to be sufficiently long for a hexagonal structure to
emerge and saturate.
If the electric field is then switched off, and the system quickly relaxes
to a metastable state which shows a residual anisotropy in the
colloidal distribution. The total length of this final relaxation phase
is $1\times10^6$ LB time steps.

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $Wa/K$\\
\hline 
Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3\\
\hline
Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3 \\
\hline
Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
\hline
External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
\hline
\end{tabular}
\end{center}
\caption{Free energy parameters used in the simulations. Bulk BPI
correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
geometries are shown in Fig.~2; and the external field case is
relevant for Fig.~3.}
\label{table:params}
\end{table}

@@

@@

Given these mappings, all quantities can be easily transformed from
simulation to physical units and vice versa. For example Figs.~1 and~2
use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
which translate into $0.01-0.1$~Pa~s respectively. This is sensible (if 
slightly low) for a molecular nematogen in the isotropic phase. 
The effective viscosity in ordered phases is higher, but this is ensured
in our model by the coupling to the order parameter.
Note that the density is set to unity in the simulations: this
corresponds to a fluid density which is about a thousand times larger
than in experiments. This makes no difference in practice as
the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
velocity, in our simulations around $10^{-5}$ in simulation units) 
remains small enough~\cite{codef}.

@@

\bibitem{blue1}
D.C. Wright and N.D. Mermin,
%  Crystalline liquids -- the blue phases.
{\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).

%\bibitem{marendu1}
%D. Marenduzzo, E.  Orlandini, M. E. Cates, J. M. Yeomans,
%{\rm Steady-state hydrodynamic instabilities of active liquid crystals:
%Hybrid lattice Boltzmann simulations.}
%{\it Phys. Rev. E} {\bf 76}, 031921 (2007).

%\bibitem{marendu2}
%O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates, 
%{\rm Domain growth in cholesteric blue phases:
%hybrid lattice Boltzmann simulations}.
%{\it Comput. Math. Appl.}, doi:10.1016/j.camwa.2009.08.047[dx.doi.org] (2009).

%\bibitem{marendu3}
%M. E. Cates, O. Henrich, D. Marenduzzo, K. Stratford,
%{\rm  Lattice Boltzmann simulations of liquid crystalline fluids:
%active gels and blue phases.}
%{\it Soft Matter} {\bf 5}, 3791-3800 (2009).

\bibitem{oliver1}
O. Henrich, D. Marenduzzo, K. Stratford, and M.E. Cates,
%Thermodynamics of blue phases in electric fields
\textit{Phys. Rev. E}, \textbf{81} 031706 (2010).

\bibitem{beris}
A. N. Beris, B. J. Edwards, 
{\it Thermodynamics of Flowing Systems with Internal Microstructure}
(Oxford Univ. Press, New York, 1994).

\bibitem{deGennes}
P.-G. de Gennes, J. Prost,
{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).

\bibitem{ladd94}
A.J.C. Ladd,
\textit{J. Fluid Mech.} \textbf{271}, 285 (1994); \textit{ibid.} \textbf{271}
311 (1994).

\bibitem{nguyen2002}
N.-Q. Nguyen and A.J.C. Ladd,
\textit{Phys. Rev. E} \textbf{66} 046708 (2002).

\bibitem{juho}
J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).

\bibitem{denniston}
C. Denniston, E. Orlandini, J. M. Yeomans, 
\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).

\bibitem{denniston2}
C. Denniston, D. Marenduzzo, E. Orlandini, J. M.  Yeomans, 
\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).

\bibitem{fournier2005}
J.-B. Fournier and P. Galatola,
\textit{Europhys. Lett.} \textbf{72}, 403 (2005).

\bibitem{oliver-bp3}
O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).

\bibitem{oliver2}
O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).

\bibitem{codef} M. E. Cates,  et al., 
\textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
\end{center}
\caption{Panels A and B as for Fig. 3, but with the disclination lines
removed for clarity. Panel C shows the situation when the field is
switched off: a residual anisotropic structure remains which reflects the
structure induced by the electric field. The electric field direction
when switched on (panel B) is into the plane of the paper.}
\end{figure}

>>>>

In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
free energy density is also present to represent the energetic cost of
anchoring at a solid surface. If the preferred order parameter tensor
at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
density (per unit area) is
\begin{equation}
f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
\end{equation}
where $W$ is a constant determining the strength of the anchoring.
The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
discussed below. For a colloid of radius $R$, the strength of the surface
anchoring compared which the bulk fluid elastic constant may be quantified
by the dimensionless parameter $WR/K$. For small values of this parameter,
the presence of a colloid surface should have little impact on the local
fluid LC ordering.

@@

The surface anchoring for colloids is as before: always normal, but with
$WR/K = 0.23$ and $WR/K = 2.3$. The colloid solid volume fractions are
1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
in simulation units. All simulations are run for two million simulation
time steps.

@@

For the simulations of BPIII with colloidal particles a system size of
$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
The simulations are
performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
An initial configuration of
randomly oriented and positioned double twist cylinders is used,
from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
is equilibrated for $6\times10^{5}$ LB time steps until no significant
further evolution is observed.

The colloidal particles with weak
normal anchoring ($WR / K= 0.23$), are then inserted with randomly chosen positions,
and the composite system is equilibrated for another $1.2\times10^{6}$
time steps.
At the end of the equilibration phase the uniform electric field
with reduced field strength ${\cal E} = 0.8$ 
is switched on (oriented along the $z$-direction), and the simulation
run for a further $1.2\times10^{6}$ time steps.
This is found to be sufficiently long for a hexagonal structure to
emerge and saturate.
If the electric field is then switched off, and the system quickly relaxes
to a metastable state which shows a residual anisotropy in the
colloidal distribution. The total length of this final relaxation phase
is $1\times10^6$ LB time steps.

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
\hline 
Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3\\
\hline
Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
0.23, 2.3 \\
\hline
Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
\hline
External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
\hline
\end{tabular}
\end{center}
\caption{Free energy parameters used in the simulations. Bulk BPI
correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
geometries are shown in Fig.~2; and the external field case is
relevant for Fig.~3.}
\label{table:params}
\end{table}

@@

{\bf 4.4 Computation of the structure factor}

In order to demonstrate structural information about the field-aligned 
states and the residual anisotropy after switch off we followed a 
procedure similar to the one described in \cite{oliver-bp3}.
The structure factor in the current approach is defined via
the Fourier transform of the colloid density:

\begin{eqnarray}
S({\bf k})&=&|F({\bf k})|^2\\
F({\bf k})&=&\int_V d{\bf r} \rho({\bf r}) e^{i {\bf k r}}
\end{eqnarray}

The colloid density $\rho({\bf r})$ is a discretised density and taken to be
$\rho({\bf r})=1$ if a lattice site at ${\bf r}$ is part of a colloid
and $\rho({\bf r})=0$ otherwise. This definition, which deviates from the 
standard definition of a structure factor and does not consider the colloids 
as pointlike objects, allows to gain the same structural information
in a simpler way.

@@

Given these mappings, all quantities can be easily transformed from
simulation to physical units and vice versa. For example Figs.~1 and~2
use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
which translate into $0.01-0.1$~Pa~s respectively. This is sensible (if 
slightly low) for a molecular nematogen in the isotropic phase. 
The effective viscosity in ordered phases is higher, but this is ensured
in our model by the coupling to the order parameter.
Note that the density is set to unity in the simulations: this
corresponds to a fluid density which is about a thousand times larger
than in experiments. This makes no difference in practice as
the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
velocity, in our simulations around $10^{-5}$ in lattice units, 
and $\Lambda$ is a typical length scale, e.g. the size of a unit cell) 
remains small enough~\cite{codef}.

@@

\bibitem{blue1}
D.C. Wright and N.D. Mermin,
%  Crystalline liquids -- the blue phases.
{\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).
\bibitem{deGennes}
P.-G. de Gennes, J. Prost,
{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).
%\bibitem{marendu1}
%D. Marenduzzo, E.  Orlandini, M. E. Cates, J. M. Yeomans,
%{\rm Steady-state hydrodynamic instabilities of active liquid crystals:
%Hybrid lattice Boltzmann simulations.}
%{\it Phys. Rev. E} {\bf 76}, 031921 (2007).
%\bibitem{marendu2}
%O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates, 
%{\rm Domain growth in cholesteric blue phases:
%hybrid lattice Boltzmann simulations}.
%{\it Comput. Math. Appl.}, doi:10.1016/j.camwa.2009.08.047[dx.doi.org] (2009).
%\bibitem{marendu3}
%M. E. Cates, O. Henrich, D. Marenduzzo, K. Stratford,
%{\rm  Lattice Boltzmann simulations of liquid crystalline fluids:
%active gels and blue phases.}
%{\it Soft Matter} {\bf 5}, 3791-3800 (2009).
\bibitem{oliver1}
O. Henrich, D. Marenduzzo, K. Stratford, and M.E. Cates,
%Thermodynamics of blue phases in electric fields
\textit{Phys. Rev. E}, \textbf{81} 031706 (2010).
\bibitem{beris}
A. N. Beris, B. J. Edwards, 
{\it Thermodynamics of Flowing Systems with Internal Microstructure}
(Oxford Univ. Press, New York, 1994).
\bibitem{ladd94}
A.J.C. Ladd,
\textit{J. Fluid Mech.} \textbf{271}, 285 (1994); \textit{ibid.} \textbf{271}
311 (1994).
\bibitem{nguyen2002}
N.-Q. Nguyen and A.J.C. Ladd,
\textit{Phys. Rev. E} \textbf{66} 046708 (2002).
\bibitem{juho}
J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).
\bibitem{denniston}
C. Denniston, E. Orlandini, J. M. Yeomans, 
\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
\bibitem{fournier2005}
J.-B. Fournier and P. Galatola,
\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
\bibitem{oliver-bp3}
O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
\bibitem{denniston2}
C. Denniston, D. Marenduzzo, E. Orlandini, J. M.  Yeomans, 
\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
\bibitem{oliver2}
O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
\bibitem{codef} M. E. Cates,  et al., 
\textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
\end{center}
\caption{Panels A and B as for Fig. 3, but with the disclination lines
removed for clarity. Panel C shows the situation when the field is
switched off: a residual anisotropic structure remains which reflects the
structure induced by the electric field. The electric field direction
when switched on (panel B) is into the plane of the paper and along z.}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=0.24\columnwidth]{support-fig8a.png}
\includegraphics[width=0.24\columnwidth]{support-fig8b.png}
\includegraphics[width=0.24\columnwidth]{support-fig8c.png}
\includegraphics[width=0.24\columnwidth]{support-fig8d.png}
\end{center}
\caption{Particle distribution during and after switched electric field direction: 
The above images show the ordering with the electric field along the x-direction 
after a previous run that had the field along z. For clarity disclination lines have
been removed. The left panels show the configuration and the structure factor 
while the field is on, whereas the righ panels show the same quantities after 
switch-off. The structure factor data shows wave vectors $k_y, k_z \in [-3\pi/8 , 3\pi/8]$.}
\end{figure}
